Chemical Tagging Using Minimum Message Length. I.\\ Gaussian Mixture Modelling

### 2017-04-24 articles/chemical-tagging-gmm/ms.tex
Some text about mml

% Have some maaaath.
\def\infinity{\rotatebox{90}{8}}
\def\veccov{\vect{C}}
\def\vecmean{\vect{\mu}}
\def\vectheta{\vect{\theta}}
\def\weight{w}
\def\weights{\vect{\weight}}
\def\datum{y}
\def\data{\vect{\datum}}

@@

\begin{equation}
  I(\vectheta,\data) 
    = \frac{p}{2}\log{q_p} - \log{\left(\frac{h(\vectheta)}{\sqrt{|\mathcal{F}(\vectheta)|}}\right)}
      - \mathcal{L}\left(\data|\vectheta\right) + \frac{p}{2}
\end{equation}

\noindent{}is the MML estimate.
Here $h(\vectheta)$ is the prior on the hypothesis, $p$ is the number of free
parameters in the model, and $|\mathcal{F}\left(\vectheta\right)|$ is the
determinant of the \emph{expected} Fisher information matrix (the second-order
partial derivatives of the negative log-likelihood function $-\mathcal{L}(\data|\vectheta)$.

The message length for a model includes the length (in bits of entropy)
required to transmit the model parameters, and the length required to transmit
the residuals of the data given the model.
There is a wealth of literature on the successes (and derivatives) of MML
in the fields of medicine, statistics, etc.

- Deep grounds in statistics, etc.

The MML principle requires that the prior beliefs of the model to be fully 
specified as part of the message. 
This requirement provides a familiar link to Bayesian statistics, and allows
one to think of MML as a form of a Bayesian optimization problem that is 
deeply grounded in information theory.
Unfortunately, strict MML is generally intractable, and therefore we require
an app

In order to minimize the full message length, we must specify the priors on
our parameters.

\subsection{Prior probabilities on the parameters}

We assume a uniform prior on the mixing weights $\bm{w}$, only requiring that
$\sum_{k=1}^{K}w_k = 1$. A useful extension of this work might be to impose
a Dirchlet prior on the mixing weights $\bm{w}$, or a distribution inspired by
the stellar mass function and observed selection function.

We adopt an improper uniform prior of 
$\mathcal{U}(\bm{\mu}) = [-\infinity, \infinity]$ on all multivariate mean 
abundances $\bm{\mu}$ (in all $D$ dimensions). This prior has no impact on
our inferences and becomes proper when it enters the (accepted) posterior
distribution. We similarily could adopt a proper and large uniform prior on all 
mean abundances of $\mathcal{U}(\bm{\mu}) = [-12, 12]$ and the result would be
the same.

@@

\begin{equation}
    p(\bm{\mu},\bm{\Sigma}) \propto |C|^{\frac{d + 1}{2}} \quad .
\end{equation}

@@

Computing the expected Fisher information requires the second order partial
derivatives of the negative log-likelihood $-\mathcal{L}(\bm{y}|\bm{\mu},\bm{\Sigma})$.
This can be tricky for mixtures -- explain why.
We approximate the determinant of of the Fisher information matrix,
$|\mathcal{F}(\bm{\mu},\bm{\Sigma})|$ as the product of $|\mathcal{F}\left(\bm{\mu}\right)|$ and $|\mathcal{F}\left(\bm{\Sigma}\right)|$ \citep{Oliver_1996,Roberts_1998}.

% Equation for log-likelihood of the data:
% \begin{equation}
% \mathcal{L}\left(\bm{y}|\bm{\mu},\bm{\Sigma}) = -\frac{Nd}{2}\log{\left(2\pi\right)} - \frac{N}{2}\log{|\bm{\Sigma}|} - \frac{1}{2}\Sigma_{i=1}^{N}(\bm{y}_i - \bm{mu})^\intercal{}\Sigma^{-1}(\bm{y}_i - \bm{\mu})
%\end{equation}

Taking the second derivative of the log-likelihood of the data with respect to $\bm{\mu}$:

\begin{equation}
\frac{\delta\mathcal{L}}{\delta\bm{\mu}} = \Sigma_{i=1}^{N}\bm{C}^{-1}(\bm{y}_i - \bm{\mu})
\end{equation}

\begin{equation}
\frac{\delta^2\mathcal{L}}{\delta\bm{\mu}^2} = N\bm{C}^{-1}
\end{equation}

Computing the Fisher information of the covariance matrix $|\mathcal{F}\left(\bm{C}\right)|$ is harder. For a full covariance matrix with off-diagonal terms, the total
number of free parameters required is $d(d+1)/2$.
\citep{Magnus_1988} derived an analytical expression:
% Dwyer_1967, Bozdogan_1990? Drton_2009?
% TODO: This derivation is a nightmare to find

\begin{equation}
|\mathcal{F}\left(\bm{C}\right) = N^\frac{d(d+1)}{2}2^{-d}|\bm{C}|^{-(d+1)}
\end{equation}

Giving the approximate determinant of the Fisher information $|\mathcal{F}\left(\bm{\mu},\bm{C}\right)|$ as:

\begin{eqnarray}
  |\mathcal{F}\left(\bm{\mu},\bm{C}\right)| & \approx & |\mathcal{F}\left(\bm{\mu}\right)|\cdot|\mathcal{F}\left(\bm{C}\right)| \\
  & \approx & N^{d}|\bm{C}|^{-1}N^\frac{d(d+1)}{2}2^{-d}|\bm{C}|^{-(d+1)} \\
  |\mathcal{F}\left(\bm{\mu},\bm{C}\right)| & \approx & N^\frac{d(d+3)}{2}2^{-d}|\bm{C}|^{-(d+2)}
\end{eqnarray}

A common method to estimate the parameters of a Gaussian distribution is by
maximum likelihood, where the log-likelihood $\mathcal{L}$ is maximized. In
order to do this, one computes the gradient of the log-likelihood function
with respect to the parameters and solves the resulting equations.

>>>>

% Have some maaaath.
\def\infinity{\rotatebox{90}{8}}
\def\veccov{\vect{C}}
\def\vecmean{\vect{\mu}}
\def\vectheta{\vect{\theta}}
\def\weight{w}
\def\weights{\vect{\weight}}
\def\datum{y}
\def\data{\vect{\datum}}
\def\likelihood{\mathcal{L}}

@@

\begin{equation}
  I(\vectheta,\data) 
    = \frac{p}{2}\log{q_p} - \log{\left(\frac{h(\vectheta)}{\sqrt{|\mathcal{F}(\vectheta)|}}\right)}
      - \likelihood\left(\data|\vectheta\right) + \frac{p}{2}
      \label{eq:message-length}
\end{equation}

\noindent{}is the MML estimate.
Here $h(\vectheta)$ is the prior on the hypothesis, $p$ is the number of free
parameters in the model, $q_p$ is the lattice quantization constant
\citep[e.g.,][]{Conway_1984}, and $|\mathcal{F}\left(\vectheta\right)|$ is the
determinant of the \emph{expected} Fisher information matrix (the second-order
partial derivatives of the negative log-likelihood function 
$-\likelihood(\data|\vectheta)$).

%The message length for a model includes the length (in bits of entropy)
%required to transmit the model parameters, and the length required to transmit
%the residuals of the data given the model.
%There is a wealth of literature on the successes (and derivatives) of MML
%in the fields of medicine, statistics, etc.

%- Deep grounds in statistics, etc.

The MML principle requires that the prior beliefs of the model to be fully 
specified as part of the message. 
We must explicitly specify our prior beliefs on the model parameters described
in Section \ref{sec:model} in order to minimize the full message length.
Indeed, the message must fully specify all components necessary for the
receiver to be able to reconstruct the message.
For the model specified in Section \ref{sec:model}, this requirement forces us
to specify a number of components:

\begin{enumerate}
  \item The number of Gaussian mixtures, $K$.
  \item Encoding the relative weights $\weights$ of the Gaussian mixtures.
  \item Encoding the component parameters $\vecmean$, $\veccov$ for all of
        the Gaussian mixtures.
  \item Encoding the data, given the model parameters.
  \item The lattice quantization constant $q_p$ for the number of model
        parameters $p$.
\end{enumerate}

Some of these components require non-trivial derivations or approximations in
order to calculate the length of the message required to encode the component.
For this reason, we will consider the message length of each component in turn
before returning to the full message length specified in Equation \ref{eq:message-length}.

\subsubsection{Encoding the number of Gaussian mixtures, $K$}
\label{sec:encoding-k}

We assume that fewer mixtures are more likely than a larger number of mixtures.
Specifically, we assume that $p(K) \propto 2^{-K}$, which implies that the
message length required to encode $K$ is given by:

\begin{equation}
  I(K) = -\log{p(K)} = K\log{2} + \textrm{constant} \quad .
\end{equation}

\subsubsection{Encoding the relative weights, $\weights$}
\label{sec:encoding-weights}

We assume a uniform prior on the mixing weights $\bm{w}$, only requiring that
$\sum_{k=1}^{K}w_k = 1$.
% This implies the weights are parameters of a multinomial distribution.
\citet{Boulton_1969} give the following expression for the message length 
required to encode the weights,

\begin{equation}
I(\weights) = \frac{K - 1}{2}\log{N} - \frac{1}{2}\sum_{k=1}^{K}\weight_k - \log{\Gamma{\left(K\right)}} \quad ,
\end{equation}

\noindent{}where $\Gamma(K)$ is the usual gamma function for positive integers
$\Gamma(K) = (K - 1)!$.
A useful extension of this work might be to impose a Dirchlet prior on the 
mixing weights $\weights$, or a distribution inspired by the stellar mass 
function and/or an observed selection function.

%This requirement provides a familiar link to Bayesian statistics, and allows
%one to think of MML as a form of a Bayesian optimization problem that is 
%deeply grounded in information theory, with successful contributions in
%economics, medicine, statistics, etc \citep{citations_needed}.

\subsubsection{Encoding the mixture parameters $\vecmean$ and $\veccov$}
\label{sec:encoding-mixture-parameters}

Recall that what we need to encode is both our priors on the model parameters
and the determinant of the expected fisher information matrix.
For the $k$-th mixture this becomes,

\begin{equation}
  I(\vecmean_k,\veccov_k) = -\log{\left(\frac{p(\vecmean_k,\veccov_k)}{\sqrt{|\mathcal{F}\left(\vecmean_k,\veccov_k\right)|}}\right)}
                          = -\log{p(\vecmean_k,\veccov_k)} + \frac{1}{2}\log{|\mathcal{F}\left(\vecmean_k,\veccov_k\right)|}
\end{equation}

\noindent{}and for all mixtures:

\begin{equation}
I(\vecmean,\veccov) = -\sum_{k=1}^{K}\log{p(\vecmean_k,\veccov_k)} + \frac{1}{2}\sum_{k=1}^{K}\log{|\mathcal{F}\left(\vecmean_k,\veccov_k\right)|}
\end{equation}

This requires us to explicitly specify our priors on the parameters $\vecmean$
and $\veccov$.
We adopt an improper uniform prior of 
$\mathcal{U}(\vecmean) = [-\infinity, \infinity]$ on all multivariate mean 
abundances $\vecmean$ (in all $D$ dimensions).
This prior has no impact on our inferences and becomes proper when it enters 
the (accepted) posterior distribution. 
We similarily could adopt a proper and large uniform prior on all mean 
abundances of $\mathcal{U}(\vecmean) = [-12, 12]$ (24 orders of magnitude!) 
and the result would be the same.

@@

\begin{equation}
    p(\vecmean_k,\veccov_k) \propto |\veccov_k|^{\frac{D + 1}{2}} \quad .
\end{equation}

% The fisher information matrix

\subsection{Prior probabilities on the parameters}
\label{sec:priors}

@@

Computing the expected Fisher information requires the second order partial
derivatives of the negative log-likelihood $-\likelihood(\data|\vecmean,\veccov)$.
This is difficult to calculate for mixtures of Gaussians, and for this reason
we approximate the determinant of the Fisher information 
$|\mathcal{F}(\vecmean, \veccov)|$ as the product of 
$|\mathcal{F}\left(\vecmean\right)|$ and $|\mathcal{F}\left(\veccov\right)|$ 
\citep{Oliver_1996,Roberts_1998}. 
We begin by taking second derivative of the log-likelihood function
$\likelihood\left(\data|\vecmean,\veccov\right)$,

\begin{equation}
  \likelihood\left(\data|\vecmean, \veccov\right) = 
  -\frac{Nd}{2}\log{\left(2\pi\right)} - \frac{N}{2}\log{|\veccov|} - \frac{1}{2}\Sigma_{i=1}^{N}\left(\data_i - \vecmean\right)^\intercal{}\veccov^{-1}\left(\bm{y}_i - \vecmean\right)
\end{equation}

\noindent{}with respect to $\vecmean$,

\begin{eqnarray}
  \frac{\delta\likelihood}{\delta\vecmean} = \Sigma_{i=1}^{N}\veccov^{-1}(\data_i - \vecmean)
  & \hspace{1em}\textrm{and}\hspace{1em}
  & \frac{\delta^2\likelihood}{\delta\vecmean^2} = -N\veccov^{-1} \quad ,
\end{eqnarray}

\noindent{}and for a $d$-dimensional vector $\vecmean$ the Fisher information
matrix is:

\begin{eqnarray}
  |\mathcal{F}(\vecmean)| = \left|-\frac{\delta^2\likelihood}{\delta\vecmean^2}\right| = N^d|\veccov|^{-1} \quad .
\end{eqnarray}

Computing the Fisher information of the covariance matrix 
$|\mathcal{F}\left(\veccov\right)|$ is harder. 
For a full covariance matrix with off-diagonal terms, the total number of free 
parameters required is $d(d+1)/2$.
\citep{Magnus_1988} derived an analytical expression:

% Dwyer_1967, Bozdogan_1990? Drton_2009?
% TODO: This derivation is a nightmare to find

\begin{equation}
  |\mathcal{F}\left(\veccov\right)| = N^\frac{d(d+1)}{2}2^{-d}|\veccov|^{-(d+1)}
\end{equation}

\noindent{}Giving the approximate determinant of the expected Fisher information 
$|\mathcal{F}\left(\vecmean,\veccov\right)|$ as:

\begin{eqnarray}
  |\mathcal{F}\left(\vecmean,\veccov\right)| & \approx & |\mathcal{F}\left(\vecmean\right)|\cdot|\mathcal{F}\left(\veccov\right)| \nonumber \\
  |\mathcal{F}\left(\vecmean,\veccov\right)| & \approx & N^{d}|\veccov|^{-1}N^\frac{d(d+1)}{2}2^{-d}|\veccov|^{-(d+1)} \nonumber \\
  |\mathcal{F}\left(\vecmean,\veccov\right)| & \approx & N^\frac{d(d+3)}{2}2^{-d}|\veccov|^{-(d+2)} \quad .
\end{eqnarray}

\subsection{Encoding the message}

Given our priors, likelihood function, and approximate expressions for the
Fisher information, we are now in a position to calculate the total message
length.
There are a number of components that go into the message:

A common method to estimate the parameters of a Gaussian distribution is by
maximum likelihood, where the log-likelihood $\likelihood$ is maximized. In
order to do this, one computes the gradient of the log-likelihood function
with respect to the parameters and solves the resulting equations.
